Structural Determinants of Sequence Evolution in Enzymatic proteins

### 2015-04-17 manuscript/voronoi/mnv.tex
new results, figures,...MN updated

\vfill
\Title{Predicting Sequence Variability from Voronoi Tessellation of Proteins}
\vfill
\Author{Amir Shahmoradi$^{1}$, Claus O. Wilke$^2$}
\Address{\affiliation}
\vfill
\begin{Abstract}
    What are the best structural predictors of protein's sequence evolution? A number of site-specific structural properties have been proposed over the past decade to answer this question. The majority of these quantities however, depend on the set of atomic coordinates used to represent individual sites in proteins, and often involve one or more number of adjustable parameters in their definition. A number of studies have already demonstrated that the choice of $C_\alpha$ atomic coordinates may not be an optimal representation of the protein's 3-dimensional structure, in particular for the calculation of site-specific quantities such as Weighted Contact Number.  Expanding on these studies and using a dataset of 209 monomeric enzymes, here we propose a new set of parameter-free structural variables derived from the Voronoi tessellation of protein structure which perform equally well or better than virtually all previously-considered structural quantities in predicting protein sequence evolution. We further show that the ideal representation of the 3-dimensional structure of proteins is the set of geometric average coordinates of atoms in the side chains of individual amino acids versus the common choice of backbone $C_\alpha$ coordinates.
    % We finally argue and show that there is no uniquely-defined best kernel for the calculation of the Weighted Contact Number, which is commonly defined as the sum of the inverse-square of the reciprocal distances between pairs of sites in proteins. By contrast, other definitions perform equally well or better.  This finding highlights the diverse energy landscapes of proteins and the fact that no single potential-of-mean-force can uniquely describe all interactions between individual sites in proteins.
    % OLD abstract:
    %A dataset of 209 monomeric enzymes is considered here to carry out a comprehensive search for the potential structural determinants of the site-specific evolution of protein sequence. Based on Voronoi tessellation of protein structures, we define new site-specific structural properties that are on average capable of describing up to $30\%$ of sequence variability observed in the dataset. We show that the Voronoi cell area and volume outperform other structural proxy measures of site-specific sequence variability previously considered in the literature, such as the Relative Solvent Accessibility, Contact Number (CN), and measures of local flexibility such as the Debye-Waller factor.  Using a variety of atomic coordinates in the definition and calculation of structural properties, we show that the best representative set of coordinates for individual sites in proteins is the average of the side-chain atomic coordinates. This choice of coordinates for the definition of structural properties results in the best predictions of site-specific sequence variability. Example structural properties that show significant improvement in this regard include those derived from Voronoi tessellation, contact number and representative B-factors for individual sites in proteins. On average, the choice of geometrical centers of side-chains vs. the commonly chosen coordinates of the backbone atom $C_\alpha$ can result in $0.07$ to $0.12$ improvements in sequence-structure correlation strengths. We finally argue and show that there is no uniquely-defined best kernel for the calculation of the Weighted Contact Number, which is commonly defined as the sum of the inverse-square of the reciprocal distances between pairs of sites in proteins. By contrast, other definitions perform equally well or better.  This finding highlights the diverse energy landscapes of proteins and the fact that no single potential-of-mean-force can uniquely describe all interactions between individual sites in proteins.
\end{Abstract}
\vfill
\vfill
\end{titlepage}
\def\thefootnote{\fnsymbol{footnote}}
\setcounter{footnote}{0}
%

@@

    Although structural characteristics have been individually extensively studied and explored with regards to their association with sequence evolution, it is yet unknown whether these seemingly independent quantities are merely different manifestations of a more fundamental underlying characteristics of individual sites in proteins or each influence the sequence evolution independently. It is perceivable that quantities such as B factor, RSA, and CN, all serve as a proxy measures of local packing density of individual sites in proteins, or the local flexibility of individual amino acids. \citet{franzosa_structural_2009} use a variety of structural variables representing the local packing density to show that RSA is the key determinant of sequence evolution with packing density having only peripheral influence. Recently however, \citet{huang_mechanistic_2014} have argued, through an extensive mathematical formulation within the framework of Elastic Network Models, for the local packing density as the dominant factor in sequence variability patterns in contrast to RSA and local flexibility measures.

@@

    Motivated by the existing gaps in the current understanding of the role of flexibility and other structural properties on sequence-structure relations in proteins, here we propose and derive a new set of site-specific structural properties which, unlike CN and WCN, their definitions does not involve any free parameters, while performing equally well or better than all previously-considered structural quantities in predicting protein sequence evolution. This is done by employing tessellation methods from the field of computational geometry to calculate several new characteristics of sites in proteins, which can serve as proxy measures of local packing density and site-specific flexibility. Contrary to what is currently perceived about the role of flexibility in sequence variability \citep[e.g.,][]{huang_mechanistic_2014}, we show that the newly calculated flexibility measures outperform many of previously studied structural properties, such as RSA and the traditional definitions of Contact Number and the Weighted Contact Number (WCN), in predicting sequence evolution at residue level. \\

    Furthermore, for structural properties that are calculated based on a set of representative site coordinates,  we show that the choice of the geometric average of the side chain atomic coordinates instead of the traditional choice of $C_\alpha$ atomic coordinates, always results in significantly better predictions of site-specific sequence evolution. Similar improvements in correlations with different ste-specific structural properties and sequence variability measures are also observed if the average of side chain B factors, instead of $C_\alpha$ atomic B factor, is used as a proxy measure of site flexibility.

@@

\section{Methods}
\label{sec:methods}

    \subsection*{Protein Dataset}

        The entire analyses and results presented in this work are based on a dataset of $209$ monomeric enzymes \citep[e.g.,][]{yeh_site-specific_2014, echave_relationship_2014} randomly picked from the Catalytic Site Atlas $2.2.11$ \citep[][]{porter_catalytic_2004} with protein sizes in the sample ranging from $95$ to $1287$ amino acids, including representatives from all six main EC functional classes \citep[][]{webb_enzyme_1992} and domains of all main SCOP structural classes \citep[][]{murzin_scop:_1995}. To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by \citet{yeh_site-specific_2014} for each protein from the {\it Clean Uniprot} database following the ConSurf protocol \citep[][]{goldenberg_consurf-db:_2009, ashkenazy_consurf_2010}. Sequence alignments were then constructed using amino-acid sequences with MAFFT \citep[][]{katoh_mafft_2005}, specifying the auto flag to select the optimal algorithm for the given data set, and then back-translated to a codon alignment using the original nucleotide sequence data. The alignments were then used to calculate the site-specific sequence variability for each individual protein in dataset. For each structure, the respective sequence alignment and phylogenetic tree were used to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model \citep[][]{mayrose_comparison_2004}, hereafter abbreviated as {\it r4sJC}.
        In addition site-specific evolutionary rates, we also calculated the Shannon entropy ($H_i$) -- the sequence entropy \citep[][]{shenkin_information-theoretical_1991} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:

@@

        where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment. We use DSSP software \citep[][]{kabsch_dictionary_1983} for the calculation of the Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of \citet{tien_maximum_2013} to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins. The {\it ddG rate} estimates for all structures in the dataset were calculated using FoldX program \citep[c.f.,][ for details of the methodology employed]{echave_relationship_2014}. In brief, the site-specific quantity, ddG rate, is a proxy measure of the stability of the entire structure of protein upon substituting an amino acid in a given site with all other $19$ amino acids. Therefore, a low ddG rate for a given site would indicate a high chance of structure perturbation upon substitution and therefore high conservation of the specific amino acid in the site on evolutionary timescales.

@@

    \subsection*{Voronoi Tessellation}

        There is already extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its prediction from sequence \citep[e.g., ][]{richards_interpretation_1974, gerstein_volume_1994}. The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and development of empirical potentials \citep[e.g.,][]{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}. For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the space into regions, called {\it cells}, such that the cell for each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points (Figure \ref{fig:voronoi}). \\

        In the context of protein studies, the atomic coordinates of $C_\alpha$ backbone atoms have been widely used as the set of Voronoi seeds to partition the 3D structure of protein according to Voronoi tessellation. The properties of individual cells resulting from tessellation are then used to obtain a wide range of information on protein structure, energy landscape or protein--protein interactions. Here in this work, we apply the simplest and most widely used definition of Voronoi tessellation described above on a dataset of $209$  monomeric enzymes. We use VORO++ software \citep[][]{rycroft_voro++:_2009} to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset. Among the most important properties are the length of the cell edges, cell area and volume, number of faces of each cell, the cell eccentricity defined as the distance between the cell's seed and the geometrical center of the cell. In addition, for each cell we also calculate {\it sphericity}, a measure of the cell {\it compactness} defined as,

@@

        in which $V$ \& $A$ stand for the volume \& area of the cell respectively. For a perfectly spherical cell, $\Psi=1$, while it becomes zero for a 2-dimensional object that has no volume but only surface area.

    \subsection*{Eliminating Degeneracy in Structural Property Definitions}

        Depending on the choice of coordinates used, there exist degeneracy in the definition of some site-specific structural variables. For example, the quantity WCN is generally calculated from the coordinates of $C_\alpha$ atoms in the 3-dimensional structure of protein. The choice of $C_\alpha$ coordinates is however mainly driven by convenience in WCN calculation and there is no reason to believe this set of atomic coordinates are the best representatives for individual sites in proteins. Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein \citep[e.g., ][]{soyer_voronoi_2000}. More recently, \citet{marcos_too_2014} have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures.\\

        Despite the highly popular choice of $C_\alpha$ atomic B factor as a proxy measure of residue flexibility \citep[e.g.,][]{halle_flexibility_2002}, same definition degeneracy also exists on choice of atomic B factors that are used to represent site-specific flexibility. In addition to WCN and B factor, there is also ambiguity as to which set of residue atomic coordinates best represent individual sites in proteins for the generation of Voronoi polyhedra. \\

        In order to identify which set of atomic coordinates best represents individual sites for the calculation of WCN, B factor, and Voronoi cells, here we consider all possible choices of the representative set of atomic coordinates. These include the set of coordinates of all backbone atoms ($N$, $C$, $C_\alpha$, $O$) and the first heavy atom in the amino acid side chains ($C_\beta$). In addition, we calculate representative coordinates for each site in protein by averaging over the coordinates of all heavy atoms in the side chains. We also calculate a representative coordinate for each site by averaging over all heavy atom coordinates in the side chain and the backbone of the amino acid together. In rare cases where the side chain $C_\beta$ atom had not been resolved in the PDB file or the amino acid lacked $C_\beta$ (e.g., Glycine), the $C_\beta$ coordinate for the specific amino acid were replaced with the coordinate of the corresponding $C_\alpha$ atom in the same amino acid. \\

\section{Results}
\label{sec:results}

    \subsection*{Voronoi Cell Area and Volume as a Measure of Local Packing Density in Proteins}

    In order to assess the prediction power site-specific variables derived from Voronoi tessellation, we first calculate the geometric centers of all side-chains for each of the proteins in the dataset and use them as the seeds of Voronoi polyhedra. Plot {\it A} of Figure \ref{fig:best_voronoi} depicts the distributions of the Spearman's correlation coefficients of five most important Voronoi cell characteristics with site-specific evolutionary rates (ER). It should be noted that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and other Voronoi cell properties. In general, we observe that the cell surface area has the best prediction power compared to other cell characteristics, followed by the cell volume, cell eccentricity as defined in previous section, cell's total edge length, and the cell sphericitiy. The cell properties are also highly associated with each other. Although the Voronoi cell volume is the second best correlating variable with ER, it exhibits no significant independent correlation with ER once we control for the cell area with the median of the distribution centered at $\sim0.0$, as illustrated in plot {\it B} of Figure \ref{fig:best_voronoi}. On the contrary, the cell sphericity and eccentricity both exhibit median partial correlations of $\sim-0.1$ \& $\sim0.07$ with ER respectively, when the contribution from the Voronoi cell area is controlled. In conclusion, the cell area, volume, and edge length appear to almost represent the same property of the Voronoi cell. Other Voronoi cell characteristics, such as the number of vertices, faces and edges of the cell also tend to correlate weakly with sequence evolutionary rates. These cell characteristics are however, discrete (integer) quantities and in general have a limited range.

    Not shown here for brevity, we also obtain identical results to the above if we use sequence entropy as defined by Eqn. \ref{eqn:shannon} instead of sequence evolutionary rates, although sequence entropy generally results in weaker correlation strengths due to the discreteness and limited range inherent in its definition.

    \subsection*{Average Side Chain coordinates as the Best Representation of Protein 3D Structure}

        Depending on the set of atomic coordinates that represent the protein structure, there are $7$ different measures for site-specific structural properties such as Weighted Contact Number, B factor and Voronoi cell properties. Therefore, in order to eliminate redundant variables from dataset, we first compare the predictive power of different measures of residue characteristics based on the set of atomic coordinates used.
        \\

        For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms of each individual side chain results in WCN values that show the best correlation with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates. Specifically, WCN from average side chain coordinates (wcnSC) outperforms WCN based on $C_\alpha$ coordinates (wcnCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates (r4sJC) by a median Spearman correlation difference of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}).
        \\

        For the measure of local flexibility in proteins (B factor) we similarly find that among all $7$ representative measures of site B factors, the average of B factor values over all heavy atoms of each individual side chain (bfSC) results in the best correlations with other structural and sequence properties. Specifically, bfSC outperforms the commonly used $C_\alpha$ B factor (bfCA) in predicting RSA, \ddg entropy, sequence entropy and evolutionary rates by a median Spearman correlation difference of $0.11$, $0.12$, $0.08$ \& $0.09$, respectively (Figure \ref{fig:best_bf}).
        \\

        Similar to WCN and Bfactor, the Voronoi cell properties, most importantly the cell surface area, volume, edge length, eccentricity and the cell sphericity also correlate best with other structure and sequence properties, only if the geometric average of side chain coordinates are used as the seeds of Voronoi cells (Figure \ref{fig:best_voronoi}).
        \\

\section{Discussion and Concluding Remarks}
\label{sec:dcr}

		Throughout the previous sections, a new set of parameter-free site-specific structural quantities were introduced using Voronoi tessellation that are capable of explaining sequence evolutionary rates equally or better than many of the previously considered site-specific structural characteristics, including RSA, ddG rate as defined by \citet{echave_relationship_2014}, site-specific measures of flexibility such as B factor, and the traditional definition of Contact Number and the Weighted Contact Number using $C_\alpha$ atomic coordinates. It is however notable that, once we recalculate WCN using the geometric center of the side chains as the representative coordinates of individual sites, the quantity WCN still outperforms all other structural quantities, including those derived from Voronoi tessellation, in explaining site-specific evolutionary rates.

        All observations clearly demonstrate that individual sites in proteins are best represented by the average properties of the side chains of amino acids in the corresponding sites. In particular, the strength of structure and sequence correlations decrease when moving from side chain to backbone atoms. An exception to this general pattern is the correlation of the hydrogen-bond energies of the sites, which correlate more strongly with site characteristics calculated based on the backbone atoms instead of side chain.
        \\

        Based on the observations described in the previous paragraphs, we keep only variables measured from average side chain properties and coordinates throughout the rest of the analysis and omit all other similar measures that show only weaker correlations with other site-specific characteristics. The exclusion of these alternative measures results in a significant reduction in the number of pdb-level variables to be further analysed, without compromising generality and comprehensiveness of the analysis.
        \\

		One potential caveat with Voronoi tessellation is the edge effects, that is, Voronoi cells that are on the surface of the protein are open. To ensure that

>>>>

\vfill
%\Title{Predicting Sequence Variability from Voronoi Tessellation of Proteins}
\Title{Structural Determinants of Sequence Evolution in Enzymatic proteins}
\vfill
\Author{Amir Shahmoradi$^{1}$, Claus O. Wilke$^2$}
\Address{\affiliation}
\vfill
\begin{Abstract}
    What are the best structural predictors of protein's sequence evolution? A number of site-specific structural properties have been proposed over the past decade to answer this question. The majority of these quantities however, depend on the set of atomic coordinates used to represent individual sites in proteins, and often involve one or more number of adjustable parameters in their definition. A number of studies have already demonstrated that the choice of $C_\alpha$ atomic coordinates may not be an optimal representation of the protein's 3-dimensional structure, in particular for the calculation of site-specific quantities such as Weighted Contact Number.  Expanding on these studies, here we propose a new set of parameter-free structural variables derived from the Voronoi tessellation of protein structure which perform equally well or better than virtually all previously-considered structural quantities in predicting protein sequence evolution. We further show that the ideal representation of the 3-dimensional structure of proteins is the set of geometric average coordinates of atoms in the side chains of individual amino acids versus the common choice of backbone $C_\alpha$ coordinates. Long-range interactions between individual amino acids are found to have a minor but significant contribution to the evolutionary patterns of protein sequence.
    % OLD abstract:
    %A dataset of 209 monomeric enzymes is considered here to carry out a comprehensive search for the potential structural determinants of the site-specific evolution of protein sequence. Based on Voronoi tessellation of protein structures, we define new site-specific structural properties that are on average capable of describing up to $30\%$ of sequence variability observed in the dataset. We show that the Voronoi cell area and volume outperform other structural proxy measures of site-specific sequence variability previously considered in the literature, such as the Relative Solvent Accessibility, Contact Number (CN), and measures of local flexibility such as the Debye-Waller factor.  Using a variety of atomic coordinates in the definition and calculation of structural properties, we show that the best representative set of coordinates for individual sites in proteins is the average of the side-chain atomic coordinates. This choice of coordinates for the definition of structural properties results in the best predictions of site-specific sequence variability. Example structural properties that show significant improvement in this regard include those derived from Voronoi tessellation, contact number and representative B-factors for individual sites in proteins. On average, the choice of geometrical centers of side-chains vs. the commonly chosen coordinates of the backbone atom $C_\alpha$ can result in $0.07$ to $0.12$ improvements in sequence-structure correlation strengths. We finally argue and show that there is no uniquely-defined best kernel for the calculation of the Weighted Contact Number, which is commonly defined as the sum of the inverse-square of the reciprocal distances between pairs of sites in proteins. By contrast, other definitions perform equally well or better.  This finding highlights the diverse energy landscapes of proteins and the fact that no single potential-of-mean-force can uniquely describe all interactions between individual sites in proteins.
\end{Abstract}
\vfill
\vfill
\end{titlepage}
\def\thefootnote{\fnsymbol{footnote}}
\setcounter{footnote}{0}
%

@@

    Although structural characteristics have been individually extensively studied and explored with regards to their association with sequence evolution, it is yet unknown whether these seemingly independent quantities are merely different manifestations of a more fundamental underlying characteristics of individual sites in proteins or each influence the sequence evolution independently. It is perceivable that quantities such as B factor, RSA, and CN, all serve as a proxy measures of local packing density of individual sites in proteins, or the local flexibility of individual amino acids. \citet{franzosa_structural_2009} use a variety of structural variables representing the local packing density to show that RSA is the key determinant of sequence evolution with packing density having only peripheral influence. Recently however, \citet{huang_mechanistic_2014} have argued, through an extensive mathematical formulation within the framework of Elastic Network Models, for the local packing density as the dominant factor in sequence variability patterns in contrast to RSA and local flexibility measures. \\

@@

    Motivated by the existing gaps in the current understanding of the role of flexibility and other structural properties on sequence-structure relations in proteins, here we propose and derive a new set of site-specific structural properties which, unlike CN and WCN, their definitions does not involve any free parameters, while performing equally well or better than all previously-considered structural quantities in predicting protein sequence evolution. This is done by employing tessellation methods from the field of computational geometry to calculate several new characteristics of sites in proteins, which can serve as proxy measures of local packing density and site-specific flexibility. Contrary to what is currently perceived about the role of flexibility in sequence variability \citep[e.g.,][]{huang_mechanistic_2014}, we show that the newly calculated flexibility measures outperform many of previously studied structural properties, such as RSA and the traditional definitions of Contact Number (CN) and the Weighted Contact Number (WCN), in predicting sequence evolution at residue level. \\

    Furthermore, for structural properties that are calculated based on a set of representative site coordinates,  we show that the choice of the geometric average of the side chain atomic coordinates instead of the traditional choice of $C_\alpha$ atomic coordinates, always results in significantly better predictions of site-specific sequence evolution. Similar improvements in correlations with different ste-specific structural properties and sequence variability measures are also observed if the average of side chain B factors, instead of $C_\alpha$ atomic B factor, is used as a proxy measure of site flexibility. \\

@@

\section{Protein Dataset and Structure$/$Sequence Variability Measures}
\label{sec:methods}

    The entire analyses and results presented in this work are based on a dataset of $209$ monomeric enzymes \citep[e.g.,][]{yeh_site-specific_2014, echave_relationship_2014} randomly picked from the Catalytic Site Atlas $2.2.11$ \citep[][]{porter_catalytic_2004} with protein sizes in the sample ranging from $95$ to $1287$ amino acids, including representatives from all six main EC functional classes \citep[][]{webb_enzyme_1992} and domains of all main SCOP structural classes \citep[][]{murzin_scop:_1995}. To assess the evolutionary rates at the amino acid level for each protein, first a set of up to $300$ homologous sequences were collected by \citet{yeh_site-specific_2014} for each protein from the {\it Clean Uniprot} database following the ConSurf protocol \citep[][]{goldenberg_consurf-db:_2009, ashkenazy_consurf_2010}. Sequence alignments were then constructed using amino-acid sequences with MAFFT \citep[][]{katoh_mafft_2005}, specifying the auto flag to select the optimal algorithm for the given data set, and then back-translated to a codon alignment using the original nucleotide sequence data. The alignments were then used to calculate the site-specific sequence variability for each individual protein in dataset. For each structure, the respective sequence alignment and phylogenetic tree were used to infer site-specific substitution rates with Rate4Site, using the empirical Bayesian method and the amino-acid Jukes-Cantor mutational model \citep[][]{mayrose_comparison_2004}, hereafter abbreviated as {\it r4sJC}. \\

    In addition site-specific evolutionary rates, we also calculated the Shannon entropy ($H_i$) -- the sequence entropy \citep[][]{shenkin_information-theoretical_1991} -- at each alignment column $i$, based on the assumption that the occurrence of each of the $20$ amino acids is equally likely at any given site in the alignments:

@@

    where $P_{ij}$ is the relative frequency of amino acid $j$ at position $i$ in the alignment. We use DSSP software \citep[][]{kabsch_dictionary_1983} for the calculation of the Accessible Surface Area (ASA) for each site normalized by the theoretical maximum solvent accessibility values of \citet{tien_maximum_2013} to obtain the Relative Solvent Accessibility (RSA) for all individual sites in all proteins. The {\it ddG rate} estimates for all structures in the dataset were calculated using FoldX program \citep[c.f.,][ for details of the methodology employed]{echave_relationship_2014}. In brief, the site-specific quantity, ddG rate, is a proxy measure of the stability of the entire structure of protein upon substituting an amino acid in a given site with all other $19$ amino acids. Therefore, a low ddG rate for a given site would indicate a high chance of structure perturbation upon substitution and therefore high conservation of the specific amino acid in the site on evolutionary timescales. \\

    A measure of thermodynamic stability changes due to amino acid substitutions at individual sites in proteins can be defined and obtained following the stability threshold model of Bloom et al. (2006) \cite{bloom_structural_2006}, which was also recently further studied by Echave et al. (2014) \cite{echave_relationship_2014}. Based upon this model which was extensively described in Chapter \ref{chap:sschar}, a quantity $\Delta\Delta G$ {\it rate} (or {it ddG rate}) was derived for each individual site in all proteins in dataset. A high ddG rate for the $i^{th}$ site in a protein indicates a high stability of the site and the overall conformation of the protein to perturbations caused by substitution of the amino acid residing the site. \\

    As a measure of local flexibility or fluctuation in different parts of the protein structure the temperature, factors ({\it B factor}) for all atoms in PDB files were extracted (c.f., Chapter \ref{chap:sschar}). Although, B factor is an atomic measure of flexibility and fluctuation in proteins, the backbone C$_\alpha$ B factor has become a very popular proxy measure of amino acid flexibility in the studies of protein dynamics and benchmarking of different Elastic Network Models of proteins.  Alternatively, the site-specific fluctuation could be calculated from MD simulations. This was however impossible for this study for the large dataset of $209$ proteins considered in this work. \\

@@

\section{Results}

    \subsection*{Voronoi Partitioning of Protein's Structure}
    \label{sec:voro}

    There is already extensive body of literature on the applications of different methods of structural partitioning in the studies of protein structure and its prediction from sequence \cite{richards_interpretation_1974, gerstein_volume_1994}. The Voronoi tessellation and its dual graph, the Delaunay triangulation, have particularly attracted much attention in the studies of protein internal structure and development of empirical potentials \cite{zomorodian_geometric_2006, zhou_alpha_2014, xia_identifying_2014}. For a given a set of centroid points (seeds) in 3-dimensional Euclidean space, the simplest and most familiar case of Voronoi tessellation divides the space into regions, called {\it cells}, such that the cell for each centroid point consists of every region in space whose distance is less than or equal to its distance to any other centroid points (Figure \ref{fig:voronoi}). \\

    % Figure goes here xx

    In the context of protein studies, the atomic coordinates of $C_\alpha$ backbone atoms have been widely used as the set of Voronoi seeds to partition the 3D structure of protein according to Voronoi tessellation. An example of Voronoi tessellation of protein structure in two dimensions (PDB ID: {\it 1LBA}) is shown in Figure \ref{fig:voronoi}.  The properties of individual cells resulting from tessellation are then used to obtain a wide range of information on protein structure, energy landscape or protein--protein interactions. \\

    Here in this work, the simplest and most widely used definition of Voronoi tessellation described above is applied on a dataset of $209$  monomeric enzymes. We use VORO++ software \cite{rycroft_voro++:_2009} to calculate the relevant Voronoi cell properties of all sites in all proteins in the dataset. Among the most important properties are the length of the cell edges, cell area and volume, number of faces of each cell, the cell eccentricity defined as the distance between the cell's seed and the geometrical center of the cell. A measure of the cell {\it eccentricity} can be also obtained by finding the distance between the cell seed and geometrical center of the cell. In addition, the cell {\it sphericity} can be calculated as a measure of the cell's {\it compactness} defined as,

@@

    \noindent in which $V$ \& $A$ stand for the volume \& area of the cell respectively. For a perfectly spherical cell, $\Psi=1$, while it becomes zero for a 2-dimensional object that has no volume but only surface area.

    \subsection*{Voronoi Cell Area and Volume as Proxy Measures of Local Packing Density and Flexibility in Proteins}

    % Figure goes here xx

    % Figure goes here xx

    In order to assess the prediction power of site-specific variables derived from Voronoi tessellation, first the geometric centers of all side-chains for each of the proteins in dataset were calculated and used as the seeds of Voronoi polyhedra. Figure \ref{fig:voronoi_ER} depicts the distributions of the Spearman's correlation coefficients of five most important Voronoi cell characteristics with site-specific evolutionary rates (ER). It is notable that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and other Voronoi cell properties. In general, it is observed that the cell surface area has the best prediction power compared to other cell characteristics, followed by the cell volume, cell eccentricity as defined in previous section, cell's total edge length, and the cell sphericitiy. \\

    The cell properties are also strongly correlated with each other. Although the Voronoi cell volume is the second best correlating variable with ER, it exhibits no significant independent correlation with ER once we control for the cell area, with the median of its distribution centered at $\sim0.0$, as illustrated in Figure \ref{fig:voronoi_ER_cond}. Conversely, the cell sphericity and eccentricity both exhibit median partial correlations of $\sim-0.1$ \& $\sim0.07$ with ER respectively, when the contribution from the Voronoi cell area is controlled. In conclusion, the cell area, volume, and edge length appear to almost represent the same property of the Voronoi cell. Other Voronoi cell characteristics, such as the number of vertices, faces and edges of the cell also tend to correlate weakly with sequence evolutionary rates. These cell characteristics are however, discrete (integer) quantities and in general have a limited range. \\

    Not shown here for brevity, almost identical results to the above are obtained if sequence entropy as defined by Eqn. \ref{eqn:shannon} were used in place of sequence evolutionary rates. The use of sequence entropy however, generally results in weaker correlation strengths due to the discreteness and limited range inherent in the definition of sequence entropy. \\

    One potential caveat with Voronoi tessellation of finite structures in Euclidean space is the {\it edge effects}. Sites that are close to the surface of protein are often associated with Voronoi cells that are bounded by the cubic box containing the protein (Figure \ref{fig:voronoi}). Here to ensure that these edge effects do not influence the observed sequence-structure correlations, the open cells -- i.e., cells that are partially bounded and closed by the cubic box containing the protein -- are identified in all proteins by examining the variations in individual cell volumes upon changing the size of the cubic box containing the protein to a given extreme value. The open cells in individual proteins are then ranked by the fraction of volume changes observed upon changing the box size and then normalized to the the largest volume observed among closed cells. It should be noted that the specific extreme value chosen for the box sizes of the proteins or the rank ordering of the open cells does not have any influence on the resulting correlation strengths, since the Spearman's $\rho$ by its definition is a rank correlation coefficient. \\

    Our conclusion is that the {\it edge effects} due to Voronoi tessellation appear to have $\lesssim0.01$ influence on the observed sequence-structure correlations in the dataset of $209$ proteins considered in this work. Similar conclusions are reached if the open cells were alternatively ranked by different criteria such as the fractional changes in cell area (vs. cell volume) upon changing the box size. The Voronoi cell characteristics, in particular cell volume and cell area can be safely used in predicting sequence variability without recourse to corrections for edge effects. An exception however is cell sphericity as defined in Eqn. \ref{eqn:sphericity}, which turns out to behave differently for open and closed cells. This is well illustrated in the adjacent averaging plots of Figure \ref{fig:zr4s_JC_voro} in which the behavior open and closed Cell characteristics, averaged over all sites in all proteins in our dataset, are plotted against the $normalized$ sequence evolutionary rates. For comparison, Figure \ref{fig:zr4s_JC_nonvoro} depicts the general behavior of the normalized site-specific evolutionary rates versus site-specific sequence entropy, $\Delta\Delta G$ rate, RSA, WCN, average Side-Chain B factor, Hydrogen bond strengths. \\

    % Figure goes here xx

    % Figure goes here xx

    \subsection{Average Side-Chain Coordinates as the Best Representation of Protein 3D Structure}
    \label{sec:voronoi_SC}

    Depending on the choice of the Cartesian coordinates used, there exist degeneracy in the definition of some site-specific structural variables. For example, the quantity WCN is generally calculated from the coordinates of $C_\alpha$ atoms in the 3-dimensional structure of protein. The choice of $C_\alpha$ coordinates is however mainly driven by convenience in WCN calculation and there is no reason to believe this set of atomic coordinates is the best representative of individual sites in proteins. Indeed, some earlier works have already suggested the use of center-of-mass of side chain coordinates to represent the 3D structure of protein \cite{soyer_voronoi_2000}. More recently, Marcos \& Echave (2014)\cite{marcos_too_2014} have also shown that WCN calculated from side-chain center-of-mass coordinates generally result in significantly better correlations of WCN with sequence variability measures. \\

    Despite the highly popular choice of $C_\alpha$ atomic B factor as a proxy measure of residue flexibility \cite{halle_flexibility_2002}, same definition degeneracy also exists on choice of atomic B factors that are used to represent site-specific flexibility. In addition to WCN and B factor, there is also ambiguity as to which set of residue atomic coordinates best represent individual sites in proteins for the generation of Voronoi polyhedra. \\

    Here in this work, all possible choices of the representative set of atomic coordinates are considered in order to identify which set of atomic coordinates best represents individual sites for the calculation of WCN, B factor, and Voronoi cells. Depending on the set of atomic coordinates that represent the protein structure, there are at least $7$ different measures of each individual site-specific structural properties, such as the Weighted Contact Number, B factor and Voronoi cell properties. These include the set of coordinates of all backbone atoms ($N$, $C$, $C_\alpha$, $O$) and the first heavy atom in the amino acid side chains ($C_\beta$). In addition, representative coordinates for each site in protein are calculated by averaging over the coordinates of all heavy atoms in the side chains. Also calculated is a representative coordinate for each site by averaging over all heavy atom coordinates in the side chain and the backbone of the amino acid together. In rare cases where the side chain $C_\beta$ atom had not been resolved in the PDB file or the amino acid lacked $C_\beta$ (e.g., Glycine), the $C_\beta$ coordinate for the specific amino acid were replaced with the coordinate of the corresponding $C_\alpha$ atom in the same amino acid. The resulting Spearman's correlation strengths of site-specific evolutionary rates, sequence entropy, \ddg rate, Relative Solvent Accessibility (RSA), amino acid hydrophobicity, and Hydrogen bond energy with different measures of WCN, B factor, and Voronoi cell area are depicted in the plots of Figures \ref{fig:best_wcn}, \ref{fig:best_bf}, and \ref{fig:best_voronoi} respectively, for different sets of atomic coordinates used in the calculations. The hydrophobicity scales of amino acids residing in individual sites in proteins were taken from \cite{hessa_recognition_2005}. Other hydropobicity scales were also considered \cite{wimley_experimentally_1996, kyte_simple_1982}, however similar results are obtained for all. \\

    % Figure goes here xx

    % Figure goes here xx

    For the measure of local packing density in proteins (the Weighted Contact Number) we find that among all possible set of coordinates, the average over coordinates of all heavy atoms of each individual side chain results in WCN values that show the strongest correlation strength with other structural and sequence properties, such as RSA, Voronoi cell properties, sequence entropy, and evolutionary rates. Specifically, WCN from average side chain coordinates outperforms WCN based on $C_\alpha$ coordinates in predicting RSA, \ddg rate, sequence entropy and evolutionary rates with median Spearman correlation differences of $0.09$, $0.10$, $0.07$ \& $0.08$, respectively (Figure \ref{fig:best_wcn}). \\

    Similar to WCN, the Voronoi cell properties, most importantly the cell surface area, volume, edge length, eccentricity and the cell sphericity also correlate best with other structure and sequence properties, only if the geometric average of side chain coordinates are used as the seeds of Voronoi cells. Specifically, cell area from average side chain coordinates outperforms cell area based on $C_\alpha$ coordinates in predicting RSA, \ddg rate, sequence entropy and evolutionary rates with median Spearman correlation differences of $0.04$, $0.06$, $0.04$ \& $0.04$, respectively (Figure \ref{fig:best_voronoi}). \\

    It is notable that the standard deviations of the difference distributions for both quantities: WCN, and Voronoi cell area, are an order of magnitude smaller than the observed differences, implying that the correlation coefficients for all proteins in dataset uniformly translate to higher values by moving from C$_\alpha$ atomic coordinates to the geometric centers of the side chains, regardless of the strength of the correlation coefficients. \\
